### 2018-07-03 report.tex
Bibliography added
Extended chapter 2

%\usepackage{amsmath}
%\usepackage[format=plain]{caption}
\usepackage{subcaption}
\usepackage{float} % force float
\usepackage[nottoc,numbib]{tocbibind} % make bibliography appear in toc
%\usepackage[margin=-50pt]{caption}
%\usepackage[colorinlistoftodos]{todonotes}
\usepackage[colorlinks=true, allcolors=blue]{hyperref}
\usepackage{cleveref} % for enumerate/itemize item referencing

@@

\subsection{Data generation}

\subsection{Data loading}

\subsection{Database search}

@@

@@

@@

@@

%\section{Overview of proteomics analysis software}
%
%\subsection{MaxQuant}
%
%One of the most used program for the quantification of label free proteins is the MaxQuant suite \cite{Cox2008}. It can be downloaded for free and is developed by the Max Planck institute in Germany. It consists of a user friendly GUI that provides the most needed steps in a proteomics pipeline.
%
%MaxQuant has been successfully adopted by the scientific community due to its ease of use and a comprehensive pipeline. A Google group and a tutorial are available  \footnote{\href{http://www.coxdocs.org/doku.php?id=maxquant:viewer:tutorial}{http://www.coxdocs.org/doku.php?id=maxquant:viewer:tutorial}} \footnote{\href{https://www.youtube.com/watch?v=\_AJHFHi5CxM}{https://www.youtube.com/watch?v=\_AJHFHi5CxM}}.
%
%The main processing steps incorporated in MaxQuant consist of:
%
%\begin{itemize}
%
%\item Read MS spectra files in the .RAW format, the closed format produced by ThermoScientific MS analyzers.
%
%\item Contains the Andromeda search engine \cite{Cox2011} for matching of \ac{MS2} spectra to a proteome database.
%
%\item Andromeda supports configuring aminoacid modifications and decoy search for FDR (false discovery rate) estimation.
%
%\item Increase peptide identifications by performing match between runs (MBR), which aims at transferring peptide to spectrum matches (PSMs) across replicate runs, based on precursor mass and retention times.
%
%\item Filters Andromeda results to provide a list of inferred peptides and proteins with a cutoff FDR value.
%
%\item Supports label-based and label-free quantification.
%
%\item Its output can be passed to the Perseus software for data visualization. %CITE
%
%\end{itemize}
%
%However, MaxQuant also suffers from some problems:
%
%\begin{enumerate}
%
%\item MaxQuant only runs on Microsoft Windows, impeding its integration in automated pipelines on cluster environments \cite{Argentini2016}. The processing steps cannot easily be fine tuned or exchanged with other pipelines. For example, only the Andromeda search engine is supported. This is a serious drawback, as the integration of results from several search engines, like MSGF+ \cite{Kim2014} or MS-Amanda \cite{Dorfer2014}  has been shown to further improve results \cite{Shteynberg2013}.
%
%\item It does not provide a command line interface (CLI), thus all analyses must be performed through the GUI. This hinders reproducibility and scalability.
%
%\item While it is supported in Linux through Mono, the user experience is best in Windows. As Linux is by far the most used OS in Bioinformatics, this implies that an additional OS is required to get the best experience with MaxQuant.
%
%\item Lacks a well written official documentation, as most information is made available only through talks published on Youtube or third party tutorials.
%
%\end{enumerate}
%
%\subsection{Compomics group}
%
%The Compomics group at Ghent University \footnote{\href{https://compomics.com/}{https://compomics.com/}} provides high quality  analysis software for the identification and integration of results through the SearchGUI, PeptideShaker and moFF (modest Feature Finder) programs \cite{Barsnes2018} \cite{Vaudel2015} \cite{Argentini2016}.
%
%\begin{itemize}
%
%\item SearchGUI provides a common interface to a range of search engines so that multiple engines can be used in a straightforward manner.
%
%\item PeptideShaker reads the SearchGUI output and performs quality control, gene set enrichment analysis (GSEA), and implements multiple validation filters to provide the best results.
%
%\item moFF reads from PeptideShaker output and implements a match between runs (MBR) model analogous to that in MaxQuant, to increase the number of spectra that are matched to a peptide. Furthermore, it summarises the \ac{MS1} peaks into refined features, making downstream analyses more sensitive.
%
%\end{itemize}
%
%SearchGUI and PeptideShaker are very well integrated and documented \footnote{\href{https://compomics.com/bioinformatics-for-proteomics/}{https://compomics.com/bioinformatics-for-proteomics/}}. Both provide not just beautiful GUIs, but also a comprehensive and extremely well documented CLI \footnote{\href{https://github.com/compomics/searchgui/wiki/SearchCLI}{https://github.com/compomics/searchgui/wiki/SearchCLI}} \footnote{\href{https://github.com/compomics/peptide-shaker/wiki/PeptideShakerCLI}{https://github.com/compomics/peptide-shaker/wiki/PeptideShakerCLI}}. They easily enable the development of analytical pipelines finely fitted to each use case.
%
%moFF is less documented and integrated in the workflow. Furthermore, it does not offer a GUI yet. The group is currently working to better integrate moFF and PeptideShaker \footnote{\href{https://groups.google.com/forum/\#!topic/peptide-shaker/Lqe7lYKLcHI}{https://groups.google.com/forum/\#!topic/peptide-shaker/Lqe7lYKLcHI}}.
%
%The main issue in the Compomics suite of tools is the lack of a robust label-free quantification step, as they currently recommend the export of results to a third party tool. Only the spectral count based methods emPAi (Exponentially Modified Protein Abundance Index) and NSAF (Normalized Spectral Abundance Factor), which are not robust \cite{Griffin2010}, are supported.
%
%
%\subsection{Other suites}
%
%Other software suites like ProteomeDiscoverer, Progenesis QI, GeneData Expressionist, etc, are available. They are not open, requiring a license to function. Furthermore, they are GUI oriented and make use of different data formats thus, making exchange across pipelines extremely difficult. The OpenMS \cite{Sturm2008} and Trans Proteomic Pipeline (TPP) \cite{Deutsch2011} suites are open, but they also suffer from data exchangeability and lack comprehensive documentations.
%
%\section{Label-free quantification}
%
%\subsection{MaxLFQ}
%
%The label-free quantification engine implemented in MaxQuant is MaxLFQ \cite{Cox2014}. It performs 2 minimisation steps to infer protein quantities from extracted ion currents (XIC), as stored in the spectra files.
%
%\subsubsection{Fractionated XIC aggreagation}
%
%First, due to the prefrationation of samples during the upstream mass spectrometry (MS) analysis, XIC signals for each peptide are distributed along multiple fractions. MaxLFQ summarises the XIC from several fractions for each peptide and sample into a single peptide intensity by minimising the sum of the square of the logarithm of the intensity 
% between samples via Levenberg–Marquardt optimization (see figure \ref{figure:xic_normalization} for a graphical explanation). 
%
%The intensity of peptide $P$ in sample $A$ is defined as a weighted average of $P$ \textquotesingle s XIC signals across fractions $\{1..k\}$, where the weights are the normalization factors that the minimisation algorithm seeks to find.
%%  this commented blank line prevents start of a new paragraph
%\begin{equation}\label{eq:peptide_intensity}
%I_{P,A}(\text{N}) = \sum\limits_{j=1}^{k}{\text{N}_{P,A,j} \times \text{XIC}_{P,A,j}}
%\end{equation}
%
%
%The minimisation assumes that the best normalization factors minimize the sum of squared logarithm pairwise-ratios $H$.
%%  this commented blank line prevents start of a new paragraph
%\begin{equation}\label{eq:sum_norm_intensity}
%H_P(N) = \sum\limits_{a,b} \abs{\frac{I_{P,a}(N)}{I_{P,b}(N)}}^2
%\end{equation}
%%  this commented blank line prevents start of a new paragraph
%\begin{equation}
%H(N) = \sum\limits_{p} H_p(N) 
%\end{equation}
%
%
%where $a,b$ iterates over all possible pair-wise combinations of samples where P was detected.
%
%The peptide intensities can be computed from the estimated normalization factors and the observed XIC signals.
%
%\begin{figure}[!h]
%\centering
%\includegraphics[width=0.8\textwidth]{xic_normalization}
%\caption{Taken from \cite{Cox2014}}
%\label{figure:xic_normalization}
%\end{figure}
%
%
%\subsubsection{Protein intensity inference}
%
%Second, once the peptide intensities have been summarised, the intensity of the proteins that the peptides originated from are inferred. This is achieved in three steps (see figure \ref{figure:protein_quant_inference} for a graphical explanation).
%
%\begin{enumerate}
%\item Peptide intensity ratios are computed for all possible pairwise combinations by dividing the normalized intensities obtained in the previous step.
%
%\item For every inferred protein A, its protein ratio $r_{A, a,b}$ for the pair of samples a,b is set to the median intensity of its children peptide ratio in samples a,b. The median is selected as a summarising statistic to protect from outliers. A minimal number of non-zero intensity peptides are required for the median to be valid, usually 2. Otherwise, the protein ratio for the pair of samples is set to 0.
%
%\item The correct protein intensities are assumed to minimise the sum of the squared difference of the logarithm of the protein ratios and logarithm of the protein intensities $I$.
%
%\end{enumerate}
%
%\begin{equation}\label{eq:least_squares}
%\sum_{a,b} (\log{(r_{A, a,b})} - \log{(I_{A,a})} + \log{(I_{A,b})})^2
%\end{equation}
%
%\begin{figure}[!h]
%\centering
%\includegraphics[width=0.8\textwidth]{protein_quant_inference}
%\caption{Taken from \cite{Cox2014}}
%\label{figure:protein_quant_inference}
%\end{figure}
%
%For the three most intense tryptic peptides, the signal per mole of protein was shown to be constant within a coefficient of variation of $\pm 10 \%$ \cite{Mikalsen2007}. Thus, protein intensities are linearly proportional to protein quantities and protein quantity ratios can be computed from protein intensity ratios. These values are reported in the MaxQuant output in the \texttt{proteinGroups.txt} file under columns named \textit{LFQ Intensity}.
%		
%\subsubsection{Drawbacks of the model}
%
%The LFQ model provides a powerful method to infer protein quantities from a label-free experiment. However, it makes use of point estimates in different steps of the process, namely (1) during peptide intensity aggregation, where the Levenberg-Marquardt minimum is selected (2) during protein ratio estimation, where the peptide median is selected and (3) during protein intensity estimation, where the least squares minimum is selected. These point estimation bottlenecks, specially (2), discard increasing amounts of data that could otherwise be used for more accurate results. Moreover, not just quantities, but also the uncertainty behind them, could also be provided. The only way of MaxQuant to provide uncertainty measurements is the number of peptides supporting the quantification measurement.
%
%
%\subsection{Bayesian model}
%
%Bayesian statistics provides probability measurements for observed data assuming underlying mathematical models. A model for protein quantification based on the \ac{MS1} intensities or XICs could be developed to assess both quantities and uncertainties behind the estimated quantities.
%
%While \ac{MS1} intensity measurements are dependent upon the underlying protein quantities, other factors influence the final measurement, thus adding noise and distorting the results. Besides the actual quantity, two more factors can be distinguished:
%
%\begin{itemize}
%
%\item \textbf{Sequence derived factors}: some peptides are easier to cleave for the cleaving enzyme (Trypsin, etc) than others. Moreover, the sequence of the peptide could influence the final measurement detected in the analyzer, by having different elution dynamics in the column or the analyzer.
%
%\item \textbf{Random noise}: the stochastic processes intrinsic to \ac{MS1} measurements could also have an impact.
%
%\end{itemize}
%
%The model would take as input:
%
%\begin{enumerate}
%
%\item \textbf{Features extracted from the precursor sequence}, including the surrounding aminoacids in the original protein. They can be used to model the sequenced derived factors.
%
%\item \textbf{Observed \ac{MS1} precursor intensities/XICs}.
%
%\end{enumerate}
%
%and return as output a protein quantification value for each protein, together with a measurement of the uncertainty behind it.
%
%
%
%
%\subsection{Data}
%
%The MaxLFQ paper dataset is available \footnote{\href{https://www.ebi.ac.uk/pride/archive/projects/PXD000279}{https://www.ebi.ac.uk/pride/archive/projects/PXD000279}}. The authors submitted the RAW files (>50 GB) and the search files containing the final results. They can be used to benchmark any new quantification tool or method, for example, it has been used to benchmark third party quantification tools like StPeter from the TPP \cite{Hoopmann2018}.
%
%The dataset contains the results of the quantification of proteins in 2 different proteomes (\textit{E. coli} and \textit{Homo sapiens}) from 6 samples organized in 2 conditions with 3 replicates each. In the first condition (H), both proteomes were mixed in a 1:1 ratio, while in the second condition (L), the \textit{E. coli} proteome had 3 times more contribution to the mix (3:1). Thus, the fold change of protein quantities across conditions should be 1 for human proteins and 3 for bacterial proteins.
%
%A list of the protein groups identified that could be mapped unambiguously to each species was made available \footnote{\href{http://www.mcponline.org/content/13/9/2513/suppl/DC1}{http://www.mcponline.org/content/13/9/2513/suppl/DC1}}. Thus, a dataset of thousands of proteins of known ratios between conditions, producing tens of thousands of peptides with known sequence, \ac{MS1} intensity/XIC and parent protein can be built from the publication, and used to train new quantification methods (figure \ref{fig:screenshot_peptide_data}).
%
%% latex table generated in R 3.4.3 by xtable 1.8-2 package
%% Fri Jun 01 17:11:23 2018
%\begin{table}[ht]
%\centering
%\begin{tabular}{rlrr}
%  \hline
% & Taxonomy & Peptides & Proteins \\ 
%  \hline
%1 & Escherichia coli (strain K12) & 14483 & 1556 \\ 
%  2 & Homo sapiens & 32647 & 3444 \\ 
%   \hline
%   & Total & 47130 & 5000
%\end{tabular}
%\caption{Summary table of the dataset}
%\label{table:summary_peptide_data}
%\end{table}
%
%\begin{figure}[!h]
%\includegraphics[width=1\textwidth]{dataset}
%\caption{Extract of the available dataset. Total of 47130 entries, one per peptide, for a global of 5000 proteins.}
%\label{fig:screenshot_peptide_data}
%\end{figure}
%
%This compiled dataset can be downloaded here \href{https://mega.nz/\#!AgcTgJYa!w9DoAKYRc6u-SaRy\_UIMz3aileUHXoaWgrxf-UycqiQ}{https://mega.nz/\#!AgcTgJYa!w9DoAKYRc6u-SaRy\_UIMz3aileUHXoaWgrxf-UycqiQ}
%
%\section{Extension of pipelines}
%
%The Compomics suite of programmes provides the best documented and accesible set of tools for proteomics analysis, but still lacks proper quantification tools. The implementation of a MS1 intensity-based downstream quantification tool compatible with the output of the Compomics tools would yield a free Linux compatible, CLI supported, complete proteomics pipeline featuring a robust quantification method.

>>>>

%\usepackage{amsmath}
%\usepackage[format=plain]{caption}
\usepackage{subcaption}
\usepackage{float} % force float
\usepackage[nottoc]{tocbibind}
%\usepackage[margin=-50pt]{caption}
%\usepackage[colorinlistoftodos]{todonotes}
\usepackage[colorlinks=true, allcolors=blue]{hyperref}
\usepackage{cleveref} % for enumerate/itemize item referencing

@@

\subsection{Data generation and loading}

The proteome benchmark dataset from \cite{Cox2014} was reanalysed starting at the output RAW files available at the PRIDE repository \footnote{\href{https://www.ebi.ac.uk/pride/archive/projects/PXD000279}{https://www.ebi.ac.uk/pride/archive/projects/PXD000279}}. The ThermoRawFileParser CITE THERMORAWFILEPARSER program was used to convert RAW files to the open format MGF (Mascot Generic Format) that can be used by the tools following the pipeline.

\subsection{Decoy database preparation and search}

The spectra contained in the MGF files obtained in the previous step were loaded onto the MS-GF+ search engine CITE using the SearchGUI \texttt{SearchCLI} CITE utility. The search parameters were set using the \texttt{IdentificationParametersCLI}. In order to account for potential post-translational modifications, the search was conducted allowing for the following variable modifications: oxidation of M and deamidation of N and Q. Moreover, C carbamidomethylation was set as fixed modification during the search. The enzyme was set to semispecific Trypsin, allowing for a non-tryptic cleavage on any side of the peptide. Up to two missed cleavages were allowed. The precursor tolerance was 10 ppm and the fragment tolerance 0.5 Da.

The database was created by combining the Uniprot proteomes for \textit{E. coli (strain K12)} (UP000000625) and \textit{Homo sapiens} (UP000005640), downloaded in June 2018. The decoy database was created using the \texttt{FastaCLI} utility in SearchGUI by reversing all sequences in the original database.

@@

The SearchGUI results were filtered using the default built-in checks available in the PeptideShaker utility \texttt{PeptideShakerCLI} \footnote{\href{https://github.com/compomics/peptide-shaker/issues/300}{https://github.com/compomics/peptide-shaker/issues/300}}. By default, the FDR is set to 1\%. Output was extracted via the Default PSM report, available in the \texttt{ReportCLI} utility

@@

The moFF command line utility CITE was used to perform a match-between-runs and extract apex intensity of each peak cluster. This required reinputing the original RAW files, together with the txt files produced by the Default PSM report from PeptideShaker. Output was exported to a peptide summary file, containing  one row per peptide and for every peptide, the detected apex intensity in each sample.

@@

Relative quantification was performed using the MSqRob utility by passing the peptide summary file from moFF.

@@
